\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-06-20 doc/incertae_sedis.tex
Revise first part of 2.7

\subsection{Ambiguity of imputed taxonomic characters}
It is tempting to view the matrix representation of the taxonomic tree
    as an encoding of the presence, absence, and uncertainty about
    suites of synapomorphies which allow us to diagnose a taxon.
Clearly by simply representing the taxonomic tree, we will not be
    able to recover the identity or even the number of characters that
    a taxonomist might have used when classifying the organisms in question.
Nevertheless, one might hope that a matrix representation of a taxonomy (such
    as the matrix in Figure \ref{fig:Handling-incertae-sedis-mrp} as a 
    representation of Figure \ref{fig:Handling-incertae-sedis}b) has a unique, uncontroversial 
    association with the taxonomic tree.

The rules for associated leaves in $\mathcal{I}(n)$ with state 1, leaves
     in $\mathcal{E}(n)$, and other leaves with the missing data symbol 
     do allow to encode the taxon $n$ in a taxonomic tree unambiguously.
However, these are not the only set of rules for generating a character matrix
    representation that would cause the taxonomic tree allows the characters to
    be mapped without homoplasy.
For example consider the case of the taxonomy shown in Figure \ref{fig:Placing-one-incertae}a.
Either of the two matrices shown in \ref{fig:Placing-one-incertae-MRP} could encode this
    taxonomy along with its \emph{incertae sedis} interpretation.
The matrix in Figure \ref{fig:Placing-one-incertae-MRP}a would result in the 
    taxon $V$ not being named in the supertree shown in Figure \ref{fig:Placing-one-incertae}b,
    while the matrix shown in Figure \ref{fig:Placing-one-incertae-MRP}b would lead
    to $V$ being named.
Thus, to honor the input taxonomic content accurately,
   we need more taxonomic information
    than simply a hierarchy and knowledge of which taxa can float more tipward.

Note that this example also presents a challenge for iterative supertree building and refinement
    of the taxonomy without retention of additional information.
Consider the case, in which there are two phylogenetic inputs along with the taxonomy from
    tree Figure \ref{fig:Placing-one-incertae}b with the first phylogeny, $T_1$, providing the single
    rooted split: $V1\,X \mid \bullet\,U$ and the second phylogeny, $T_2$ providing the 
    rooted split: $W\, V1 \mid \bullet\,V2$.
If the supertree pipeline were run with just the taxonomy and and $T_1$ as inputs, then the supertree
    would be the tree shown in Figure \ref{fig:Placing-one-incertae}c.
One might hope that using that tree as the revised taxonomy and then adding the split from $T_2$ in
    a subsequent analysis would result in the same tree as if $T_1$ and $T_2$ to the original taxonomy.
While the topology for the final tree in either order of analyses would indeed be the tree shown in 
    Figure \ref{fig:Placing-one-incertae}b, the decision about whether or not to name $V$ would differ.
Our handling of \emph{incertae sedis} in the original taxonomy via rooted partial splits
    would encode the taxonomic definitions in the manner depicted in Figure \ref{fig:Placing-one-incertae-MRP}a,
    but the encoding of the revised taxonomy in Figure \ref{fig:Placing-one-incertae}c would result
    in an altered definition of taxon $V$, as shown in \ref{fig:Placing-one-incertae-MRP}c.
As expected,  the definition of $WX$ has been altered to reflect the
    fact $T_1$ placed $V$ inside $WX$
    when we encode the revised taxonomy.
Perhaps, unexpected is the fact that storing the revised taxonomy with an \emph{incertae sedis} annotation
    and then reprocessing the taxonomy using our rooted partial splits approach would also shift
    the definition of $V$ from a taxon that must exclude $W$ to one that may include $W$.

@@

\caption{\label{fig:Placing-one-incertae}
One \emph{incertae sedis} group ($W$) being placed within another ($V$), with possible
loss of the taxon $V$.}
\end{figure}

>>>>

\subsection{Ambiguity of split representation for \emph{incertae sedis} taxa}
Equation (\ref{eq:exsetformonea}) introduces a method for encoding a
taxonomy with \emph{incertae sedis} taxa into a set of partial splits.
These partial splits have a unique represention as a matrix of
0-1 characters with missing data indicated by '?'.  Clearly, we will
not be able to recover the identity, or even the number, of characters
that a taxonomist might have used when classifying a set of organisms,
simply by looking at the character matrix that represents a taxonomy
(as Figure \ref{fig:Handling-incertae-sedis-mrp} represents Figure
\ref{fig:Handling-incertae-sedis}b). Nevertheless, one might hope that 
equation (\ref{eq:exsetformonea}) is a unique an uncontroversial
way to encode any taxonomic tree as a set of splits.

However, equation (\ref{eq:exsetformonea}) is not the only set of rules 
for generating a split representation for a taxonomy. For example
consider the case of the taxonomy shown in Figure
\ref{fig:Placing-one-incertae}a. Either of the two matrices shown in
\ref{fig:Placing-one-incertae-MRP}a or
\ref{fig:Placing-one-incertae-MRP}b could represent this taxonomy 
along with its \emph{incertae sedis} interpretation.  The matrix in
Figure \ref{fig:Placing-one-incertae-MRP}a, which corresponds to
equation (\ref{eq:exsetformonea}), assumes that \emph{incertae sedis}
taxa may only intrude into taxa that are siblings in the taxonomy.
The placement of $W$ within $V$ leads to taxon $V$ not being named in
the supertree shown in Figure \ref{fig:Placing-one-incertae}c under
the encoding in \ref{fig:Placing-one-incertae-MRP}a because
$V$ is not a sibling of $W$ in Figure \ref{fig:Placing-one-incertae}a.

The matrix shown in Figure \ref{fig:Placing-one-incertae-MRP}b allows
$V$ to be named in the supertree shown in Figure
\ref{fig:Placing-one-incertae}c because it removes the assumption that
\emph{incertae sedis} taxa can only be placed within siblings.
However, as a side-effect, if taxon $WX$ is found to be
non-monophyletic, the matrix from Figure \ref{fig:Placing-one-incertae-MRP}b
allows $W$ to be placed within $V$.  In general, this approach would
allow incertae sedis children of any taxon $\Gamma_1$ to be placed
arbitrary deep into group $\Gamma_2$ if group $\Gamma_1$ is broken. This
behavior can lead to surprising and non-intuitive results. Additionally, our unpruner currently assumes
that \emph{incertae sedis} taxa 
can only be placed within siblings, and therefore does not know what
to do in this situation.  Thus, it is not clear that either Figure
\ref{fig:Placing-one-incertae-MRP}a or Figure
\ref{fig:Placing-one-incertae-MRP}b comprise a unique, best
encoding of the taxonomy.  Thus, to honor the input taxonomic content
accurately, we may need more taxonomic information than simply a
hierarchy and knowledge of which taxa can float more tipward.

Note that this example also exposes a lack of consistency in
Equation (\ref{eq:exsetformonea}) that rears its head when attempting to
iteratively revise placement of \emph{incertae sedis} taxa in the taxonomy
on successive supertree builds.
% without retention of additional information.

Consider the case, in which there are two phylogenetic inputs along with the taxonomy from
    tree Figure \ref{fig:Placing-one-incertae}b with the first phylogeny, $T_1$, providing the single
    rooted split: $V1\,X \mid \bullet\,U$ and the second phylogeny, $T_2$ providing the 
    rooted split: $W\, V1 \mid \bullet\,V2$.
If the supertree pipeline were run with just the taxonomy and and $T_1$ as inputs, then the supertree
    would be the tree shown in Figure \ref{fig:Placing-one-incertae}c.
One might hope that using that tree as the revised taxonomy and then adding the split from $T_2$ in
    a subsequent analysis would result in the same tree as if $T_1$ and $T_2$ to the original taxonomy.
While the topology for the final tree in either order of analyses would indeed be the tree shown in 
    Figure \ref{fig:Placing-one-incertae}b, the decision about whether or not to name $V$ would differ.
Our handling of \emph{incertae sedis} in the original taxonomy via rooted partial splits
    would encode the taxonomic definitions in the manner depicted in Figure \ref{fig:Placing-one-incertae-MRP}a,
    but the encoding of the revised taxonomy in Figure \ref{fig:Placing-one-incertae}c would result
    in an altered definition of taxon $V$, as shown in \ref{fig:Placing-one-incertae-MRP}c.
As expected,  the definition of $WX$ has been altered to reflect the
    fact $T_1$ placed $V$ inside $WX$
    when we encode the revised taxonomy.
Perhaps, unexpected is the fact that storing the revised taxonomy with an \emph{incertae sedis} annotation
    and then reprocessing the taxonomy using our rooted partial splits approach would also shift
    the definition of $V$ from a taxon that must exclude $W$ to one that may include $W$.

@@

\caption{\label{fig:Placing-one-incertae}
  Placement of incertae sedis taxa within non-siblings could occur if
  taxa become siblings by placement.  Given the taxonomy (a),
  placement of $V$ within $WX$ is shown in (b), followed by placement
  of $W$ within $V$ as shown in (c).  Since $W$ is not a sibling of
  $V$ in (a), the method proposed in this paper and illustrated in Figure
  \ref{fig:Placing-one-incertae-MRP}a does not preserve the taxon name
  for $V$.  However, with an alternative method for computing reduced
  exclude sets that is shown in \ref{fig:Placing-one-incertae-MRP}b,
  the name $V$ would be preserved.
  }
\end{figure}
